On Optimality of Finite Element Integration

### 2015-10-24 TOMS_Paper/assembly.tex
Change again sharing elimination---ufff

@@

\begin{Def}[Outer product loop nest]
An outer product loop nest is a perfect multilinear loop nest of arity $n \leq 2$ in which all of the expressions appearing in the body of the innermost loop are summations, and every loop index occurs once in every summand.
\end{Def}

As shown later, outer product loop nests are important because they possess properties that we can take advantage of to synthesize optimal code. We will see that they arise naturally when translating bilinear or linear forms into code.

@@

\begin{Def}[Reduction loop]
A loop $L_i$ is said to be a reduction loop if in its body
\begin{enumerate}
\item $i$ appears only as an array index, and
\item for each augmented assignment statement $S$ (e.g., an increment), arrays indexed by $i$ appear only on the right hand side of $S$.
\end{enumerate}
\end{Def}

\begin{Def}[Free order loop]
A loop $L_i$ is said to be a free order loop if its iterations can be executed in any arbitrary order; that is, no loop-carried dependence exists across two different iterations. 
\end{Def}

Altogether, these definitions allow us to formalize the class of loop nests for which we seek optimality.

\begin{Def}[Finite element integration loop nest]
\label{def:fem-loopnest}
A finite element integration loop nest is a loop nest in which we identify, in order, an imperfect free order loop, a (generally) imperfect, linear or non-linear reduction loop, and an outer product loop nest.
\end{Def}

@@

To synthesize optimal implementations as in Definition~\ref{def:mln-optimality} we need:
\begin{itemize}
\item a characterization of the space of legal transformations impacting the operation count
\item a cost model to select the optimal point in the transformation space
\end{itemize}
In the following Sections~\ref{sec:sharing-elimination} and~\ref{sec:pre-evaluation}, we construct the transformation space. We leave a few loose ends, which we progressively tie up in Section~\ref{sec:optimal-synthesis}.

@@

\begin{Def}[Sharing factor]
The sharing factor of a loop $L_i$, $L^{sf}_i$, is given by the cardinality of the set of symbols depending on $i$ (i.e., arrays indexed by $i$). It provides an indication of the maximal number of symbols which can be factorized to the level of $L_i$.
\end{Def}

\begin{figure}[h]\begin{CenteredBox}
{\subfigcapskip = 7pt \subfigure[With sharing. Note that $L^{sf}_i = 2$ and $L^{sf}_j = 1$.]{\label{code:multi_loopnest_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest.code}}}
~~~~~~~~~~~~~~
{\subfigcapskip = 2pt \subfigure[Optimal form]{\label{code:multi_loopnest_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_opt.code}}}
\end{CenteredBox}\caption{Reducing a simple outer product loop nest to optimal form.}\label{code:multi_loopnest}\end{figure}

Figure~\ref{code:multi_loopnest_a} shows an example of a trivial outer product loop nest of arity $n=2$ with sharing along dimension $j$ induced by $b_j$. Figure~\ref{code:multi_loopnest_b} shows an optimal synthesis for the loop nest in Figure~\ref{code:multi_loopnest_a}. More generally, arbitrarily complex outer product loop nests can systematically be reduced to optimal form using an algorithm that exploits the properties of linearity and perfectness. The algorithm and the proof that we provide below focus on outer product loop nests of arity $n=2$. For $n < 2$ the proof would be trivial, whereas for $n > 2$ a more general treatment would be required; fortunately, this case is never of interest in finite element integration.

\begin{Algo}[Optimal code motion for outer product loop nests]
\label{algo:outer-product-optimal}
\normalfont
Consider an outer product loop nest $\Lambda = [L_i, L_j]$, with the body of $L_j$ being an independent basic block. The algorithm starts with ``flattening'' the expressions (e.g., by multiplication expansion) in the body of $L_j$; this exposes the space of all possible factorizations. Since $\Lambda$ is an outer product loop nest, the resulting expressions are summations of outer products. We traverse the expression tree and determine the sharing factor of the two linear loops. We take the symbols depending on the loop with smallest sharing factor, say $L_i$, and we incrementally factorize them. Due to linearity, each factored product $P$ only has one symbol depending on $L_i$, and this symbol is now unique in the expression. The other term in $P$, independent of $L_i$, is, by definition, loop-invariant, and therefore hoisted such that redundant computation is avoided. We repeat the same process for the symbols along the loop with higher sharing factor, $L_j$. The transformations are semantically correct: multilinearity ensures deterministic factorization; perfectness ensures safeness of hoisting.
\end{Algo}

\begin{Prop}
\label{prop:optimal-mln}
Let $\Lambda = [L_i, L_j]$ be an outer product loop nest with the body of $L_j$ being an independent basic block. Then Algorithm~\ref{algo:outer-product-optimal} leads to an optimal synthesis $\Lambda'$. 
\end{Prop}
\begin{proof}
Assume, without loss of generality, that $L_i$ is the loop with smallest sharing factor. We reduce the proof to showing that the two following facts are true (we do it by contradiction).
\begin{enumerate}
\item Taking $L_j$, the loop with higher sharing factor, as point of departure would lead to a sub-optimal $\Lambda'$ (optimal only in the best case). If we assumed this were false, there would be two possibilities to consider: A) both $L_i$ and $L_j$ present sharing over disjoint sets of outer products; B) at least one outer product induces sharing along both $L_i$ and $L_j$. Case A) is quite trivial since there are no side effects in swapping the order of factorizations, so $\Lambda'$ would actually be optimal (best case scenario). In case B), due to linearity, factorization produces a summation of $L^{sf}_{j}$ outer products, as opposed to $L^{sf}_{i}$ if we had started with $L_i$. However, note that factorization followed by hoisting is always a winning strategy: evaluation, at every iteration, of hoisted symbols requires one operation (a sum) instead of the two operations (a product and a sum) needed by an outer product. Since $L^{sf}_{j} > L^{sf}_{i}$, it is clear that $\Lambda'$ could not be optimal, which contradicts the hypothesis.
\item Partial elimination of sharing would lead to a sub-optimal $\Lambda'$ (optimal only in the best case). Let us assume, for a moment, that this is false. In this scenario, at least one symbol is excluded from the factored product. Without loss of generality, consider, for instance, $a_i(b_j + c_j + ...) + a_i e_j + ...$. We see that either there is another factorization opportunity along $L_j$ ($e_j$ is shared with some other terms), in which case no additional operations need be performed, or $\Lambda'$ is sub-optimal. Consequently, the hypothesis is generally contradicted.
\end{enumerate} 
\end{proof}

\subsection{Pre-evaluation}
\label{sec:pre-evaluation}
For finite element integration loop nests, the presence of sharing is a \textit{sufficient but not necessary} condition for being in \textit{non-optimal} form. Consider again the bilinear form implementation in Figure~\ref{code:loopnest}. We pose the following question: are we able to identify sub-expressions within $F$ for which the reduction imposed by $L_i$ can be pre-evaluated, thus obtaining a decrease in operation count proportional to the size of $L_i$, $M$? The transformation we look for is exemplified in Figures~\ref{code:loopnest_red} and~\ref{code:loopnest_nored}; Figure~\ref{code:loopnest_red}, which represents the input, can be seen as a simple instance of the abstract loop nest in Figure~\ref{code:loopnest}.

@@

Pre-evaluation opportunities can be exposed through an exploration of the expression tree transformation space. This would be challenging if we were to deal with arbitrary loop nests and expressions. However, we can make use of a result -- the foundation of tensor contraction mode -- to build a relatively simple algorithm. As summarized in Section~\ref{sec:tc}, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces; such monomials can always be reduced to a product of two tensors. This result can be turned into a loop and expression transformation algorithm, for which we provide a succinct description below.

>>>>

% line over text in non math mode
\newcommand{\textoverline}[1]{$\overline{\mbox{#1}}$}

@@

\begin{Def}[Outer product loop nest]
\label{def:jk-loop}
An outer product loop nest is a perfect multilinear loop nest of arity $n \leq 2$ in which all of the expressions appearing in the body of the innermost loop are summations, and every loop index occurs once in every summand.
\end{Def}

As shown later, outer product loop nests are important because they have a structure that we can take advantage of to synthesize optimal code. We will see that they arise naturally when translating bilinear or linear forms into code.

@@

\begin{Def}[Reduction loop]
\label{def:i-loop}
A loop $L_i$ is said to be a reduction loop if in its body
\begin{enumerate}
\item $i$ appears only as an array index, and
\item for each augmented assignment statement $S$ (e.g., an increment), arrays indexed by $i$ appear only on the right hand side of $S$.
\end{enumerate}
\end{Def}

\begin{Def}[Free order loop]
\label{def:e-loop}
A loop $L_i$ is said to be a free order loop if its iterations can be executed in any arbitrary order; that is, there are no loop-carried dependencies across different iterations. 
\end{Def}

We use these definitions to formalize the class of loop nests for which we seek optimality.

\begin{Def}[Finite element integration loop nest]
\label{def:fem-loopnest}
A finite element integration loop nest is a loop nest in which we identify, in order, an imperfect free order loop, a (generally) imperfect, linear or non-linear reduction loop, and an outer product loop nest whose body is an independent basic block.
\end{Def}

@@

To synthesize optimal implementations as in Definition~\ref{def:mln-optimality} we need:
\begin{itemize}
\item a characterization of the space of legal transformations impacting the operation count
\item a cost model to select the optimal point in the transformation space
\end{itemize}
In the following Sections~\ref{sec:sharing-elimination} and~\ref{sec:pre-evaluation}, we focus on the construction of the transformation space.

@@

\begin{figure}[h]\begin{CenteredBox}
{\subfigcapskip = 7pt \subfigure[With sharing]{\label{code:multi_loopnest_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest.code}}}
~~~~~~~~~~~~~~
{\subfigcapskip = 2pt \subfigure[Optimal form]{\label{code:multi_loopnest_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_opt.code}}}
\end{CenteredBox}\caption{Reducing a simple outer product loop nest to optimal form.}\label{code:multi_loopnest}\end{figure}

For instance, Figure~\ref{code:multi_loopnest_a} shows a trivial outer product loop nest of arity $n=2$ with sharing along dimension $j$ induced by $b_j$. The optimal synthesis for this loop nest is provided in Figure~\ref{code:multi_loopnest_b}. This example illustrates how factorization can be applied to expose code motion opportunities in multilinear loop nests.

In this section, we study an optimal transformation to perform \textit{sharing elimination}; that is, a transformation that minimizes the operation count of a loop nest by applying factorization, code motion and (potentially) common sub-expression elimination. In general, transformations of this sort require solving large combinatorial problems; for instance, to evaluate the impact of all possible factorization strategies. However, we show that by restricting ourselves to the class of loops and expressions of Definition~\ref{def:fem-loopnest}, the search space can drastically be pruned. In the following, we assume outer product loop nest of arity $n=2$ (i.e., a bilinear form), although the same reasoning stands for the case $n < 2$ (which is actually just a special case).

The first observation is that finite element expressions can always be reduced to the form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$, where $m$ depends on the specific form and discretization, $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix with test (trial) functions, and $\sigma_{ei}$ is a function of the coefficients.

Given the peculiar structure of our terms, the key idea is that, in the space $S$ of all possible factorizations, the optimum almost always lies in a tiny subspace (typically of size one). We refer to this space as the \textit{potential search space} ($PSS$). Our approach is therefore as follows: we calculate lower and upper bounds (respectively, $lb$ and $ub$) for families of factorizations and we compare them: if, as expected, $ub(PSS) < lb(\overline{PPS})$, we prune $\overline{PPS}$. Otherwise, we have to search in $\overline{PPS}$ too.

A potentially optimal factorization is one that groups together test and trial functions. Such a factorization exposes sub-expressions independent of at least one loop and, therefore, hoistable. The converse is not true: factorizing $\sigma_{ei}$ (or any of the terms in it) generally inhibits hoisting of sub-expressions including test and trial functions. Algorithm~\ref{algo:pps} provides a succinct description of how $PPS$ is built.

\begin{Algo}[Construction of the potential search space]
\label{algo:pps}
\normalfont
Consider a finite element integration loop nest $\Lambda=[L_e, L_i, L_j, L_k]$. The algorithm starts with reducing the input expression to the normal form $\sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \sigma_{ei}^{w}$.

The sub-expressions $\alpha_{eij}^{w} \beta_{eik}^{w}$ are ``flattened'' (by multiplication expansion) such that all $L_j$- and $L_k$-dependent terms lie on the same level of the expression tree. We factorize, in sequence, $L_j$- ($L_k$-) and $L_k$- ($L_j$-) dependent terms (it can easily be demonstrated, by contradiction, that one should start from the loop with the least number of unique terms in order to obtain the smallest possible operation count). Thanks to linearity, each factored product only has one symbol depending on $L_j$ ($L_k$), and this symbol is now unique in the expression. The other sub-expression, independent of $L_j$ ($L_k$), is, by definition, loop-invariant, and therefore hoisted such that redundant computation is avoided. Note that multilinearity ensures deterministic factorization, while perfectness ensures correctness of hoisting.
\end{Algo}

%A special case is when test and trial functions belong to the same function space: this induces common sub-expressions \textit{over different loops}, so a potentially winning strategy consists of hoisting both $\alpha_{eij}$ and $\beta_{eik}$ and exploiting the fact that $\alpha = \beta$. ---- If we note that, for some $w$, $\alpha = \beta$, we add the following strategy to $PPS$: hoist $\alpha_{eij}^{w}$ (or, equivalently, $\beta_{eik}^{w}$); replace common sub-expressions; minimize cost of $\sigma_{ei}$, which we discuss later.

%The loose end we have to tie up is the transformation of $\sigma_{ei}^{ab}$ terms. If $L_i$ is a linear loop, we can straightforwardly apply Algorithm~\ref{algo:outer-product-optimal} (with arity $n=1$) so that integration-dependent terms are grouped and geometry-dependent sub-expressions are hoisted. If, on the other hand, $L_i$ is non linear, we try all possible factorizations of $L_i$-dependent terms to maximize motion of geometry-dependent sub-expressions (although, we believe, a more refined approach, which avoids searching the whole space, can be found).

\subsection{Pre-evaluation}
\label{sec:pre-evaluation}
Sharing elimination aims to reduce the operation count by applying factorization, code motion and/or common sub-expression elimination. There is a fourth operator impacting the operation count: \textit{reduction pre-evaluation}. In this section, we discuss role and impact of this operator. We also show that there exists a single point ($\in \overline{PPS}$) in the whole transformation space for which reduction pre-evaluation is legal.

We start with an example. Consider again the bilinear form implementation in Figure~\ref{code:loopnest}. We pose the following question: are we able to identify sub-expressions within $F$ for which the reduction imposed by $L_i$ can be pre-evaluated, thus obtaining a decrease in operation count proportional to the size of $L_i$, $M$? The transformation we look for is exemplified in Figures~\ref{code:loopnest_red} and~\ref{code:loopnest_nored}; Figure~\ref{code:loopnest_red}, which represents the input, can be seen as a simple instance of the abstract loop nest in Figure~\ref{code:loopnest}.

@@

Pre-evaluation opportunities can be exposed through exploration of the expression tree transformation space. This would be challenging if we were to deal with arbitrary loop nests and expressions. However, we can make use of a result -- the foundation of tensor contraction mode -- to solve the problem by means of a relatively simple algorithm. As summarized in Section~\ref{sec:tc}, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces; such monomials can always be reduced to a product of two tensors. This result can be turned into a loop and expression transformation algorithm, for which we provide a succinct description below.
